### 2025-04-08 correlation.qmd
20250407 - correlation

@@

The $r$ value (`{r} pearson_r`) is slightly negative.
The [effect size](#sec-practicalSignificance) of the association is small—for what is considered a small, medium, and large effect size for $r$ values, see @sec-effectSizeThresholds.
The $p$-value is less than .05, so it is a statistically significant correlation.
We could report this finding as follows: There was a statistically significant, weak negative association between height and fantasy points ($r[$`{r} pearson_df`$] =$ `{r} pearson_r`, $p < .001$); the taller a player was, the fewer fantasy points they tended to score.

@@

Please note that the $p$-value for a correlation assumes that the observations are independent—in particular, that the residuals are not correlated.
However, the observations are not independent in the `player_stats_seasonal` dataframe used above, because the same player has multiple rows—one row corresponding to each season they played.
This non-independence violates the traditional assumptions of the significance test of a correlation.
We could address this assumption by analyzing only one season from each player.
However, for simplicity, we present results from the whole dataframe.
In @sec-mixedModels, we discuss [mixed model](#sec-mixedModels) approaches that handle repeated measures and other data that violate assumptions of non-independence that are shared by correlation and [multiple regression](#sec-multipleRegression).

>>>>

Notice the diagonal values are all 1.0.
That is because a variable is perfectly correlated with itself.
Also notice that the values above the diagonal are a mirror image of (i.e., they are the same as) the respective values below the diagonal.
The association between two variables is the same regardless of the order (i.e., which is the [predictor variable](#sec-correlationalStudy) and which is the [outcome variable](#sec-correlationalStudy)); that is, the association between variables A and B is the same as the association between variables B and A.

@@

The $r$ value (`{r} petersenlab::apa(pearson_r, digits = 2)`) is slightly negative.
The [effect size](#sec-practicalSignificance) of the association is small—for what is considered a small, medium, and large effect size for $r$ values, see @sec-effectSizeThresholds.
The $p$-value is less than .05, so it is a statistically significant correlation.
We could report this finding as follows: There was a statistically significant, weak negative association between height and fantasy points ($r[$ `{r} pearson_df` $]=$ `{r} petersenlab::apa(pearson_r, digits = 2)`, $p < .001$); the taller a player was, the fewer fantasy points they tended to score.

@@

Please note that the $p$-value for a correlation assumes that the observations are independent—in particular, that the residuals are not correlated.
However, the observations are not independent in the `player_stats_seasonal` dataframe used above, because the same player has multiple rows—one row corresponding to each season they played.
This non-independence violates the traditional assumptions of the significance test of a correlation.
We could address this assumption by analyzing only one season from each player.
However, for simplicity, we present results from the whole dataframe.
In @sec-mixedModels, we discuss [mixed model](#sec-mixedModels) approaches that handle repeated measures and other data that violate assumptions of non-independence that are shared by correlation and [multiple regression](#sec-multipleRegression); assumptions of [multiple regression](#sec-multipleRegression) are described in @#sec-assumptionsRegression.
